---
title: "Data624 Project 2"
author: "Group 4"
date: "`r Sys.Date()`"
output:
  html_document:
    font-family: Consolas
    highlight: tango
    number_sections: no
    theme: paper
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
    toc_depth: '4'
---

```{=html}
<style type="text/css">

code {
  font-family: "Consolas";
  font-size: 11px;
}

pre {
  font-family: "Consolas";
  font-size: 11px;
}

mark {
  background-color: whitesmoke;
  color: black;
}

</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=F, message=F, include = T, echo = F, fig.height=3.5)

options(scipen = 9)
set.seed(101)

library(fpp2)
library(ggplot2)
library(tidyr)
library(dplyr)
library(seasonal)
library(imputeTS)
library(inspectdf)
library(lubridate)
library(fpp3)
```

<font size="3">Group Members</font>

-   Subhalaxmi Rout
-   Kenan Sooklall
-   Devin Teran
-   Christian Thieme
-   Leo Yi

/pagebreak

## Introduction

We have been given a dataset from a beverage manufacturing company that consists of 2,571 rows of data and 33 columns. The dataset contains information on different beverages and their chemical composition. The goal of this analysis is to use the 32 predictive features to predict the *Potential for hydrogen* (pH), which is a measure of the acidity/alkalinity of the beverage. pH is the key KPI in this analysis. 

We'll begin by reading in the dataset and looking at each column's data type: 

```{r}
df <- readr::read_csv('C:/Users/chris/OneDrive/Master Of Data Science - CUNY/Summer 2021/Predictive_Analytics/Predictive-Analytics-624/Project2-PredictiveModelingNonLinearRegressionTechniques/StudentData-TO_MODEL.csv')

# convert column names to all lowercase
names(df) <- lapply(names(df), tolower)

glimpse(df)
```
We see that all columns, with the exception of `brand`, are doubles and continuous. Excluding the response variable, this means that we have 1 categorical variable and 31 continuous variables to work with. 

## Exploratory Data Analysis

In the output above, we can see that there are missing values (NAs). Let's see how pervasive this issue is within our dataset: 

```{r fig.width = 12, fig.height=6}
visdat::vis_miss(df, sort_miss = TRUE)
```

In total, only about 1% of our data is missing. We can see that most of the columns are only missing a negligible amount of data. `mfr` and `brand code` have the largest amount of missing values and are missing 8.25% and 4.67% of their data, respectively. Now that we understand that our missing values are not a pervasive issue, we'll continue with our analysis.

#### Distribution of Response Variable: pH

Let's get an understanding of the distribution of our response variable: 

```{r}
df %>% 
  select(ph) %>%
  ggplot() + 
  aes(x = ph) + 
  geom_histogram()
```

The distribution of pH is left-skewed and multi-modal. Generally speaking, when we see a multi-modal distribution, often times that is an indication that there are sub-populations within the distribution. We know from looking at our dataset that there is a `brand code` with values A, B, C, and D. Let's break up the above distribution into 4 distributions based on these values: 

```{r}
a <- df %>% 
  filter(`brand code` == 'A') %>% 
  select(ph) %>%
  ggplot() + 
  aes(x = ph) + 
  geom_histogram() +
  labs(title = "Brand A")

b <- df %>% 
  filter(`brand code` == 'B') %>% 
  select(ph) %>%
  ggplot() + 
  aes(x = ph) + 
  geom_histogram()+
  labs(title = "Brand B")

c <- df %>% 
  filter(`brand code` == 'C') %>% 
  select(ph) %>%
  ggplot() + 
  aes(x = ph) + 
  geom_histogram()+
  labs(title = "Brand C")

d <- df %>% 
  filter(`brand code` == 'D') %>% 
  select(ph) %>%
  ggplot() + 
  aes(x = ph) + 
  geom_histogram()+
  labs(title = "Brand D")

gridExtra::grid.arrange(a, b, c, d, nrow = 2)
```

Breaking down to this further grain does not seem to be much more helpful. There may be even more granular sub-populations within this data that we are not seeing. 

Now that we've looked at our response variable, let's look at our numeric explanatory variables.

```{r fig.height=15, fig.width=10}
inspectdf::inspect_num(df %>% select(-ph)) %>% 
  show_plot()
```

We note the following about these distributions: 

* `air pressurer` - there appears to be either two distributions here, or a single distribution with a pocket of outliers
* `balling`, `balling lvl`, `density`,`fill pressure`, `hyd pressure1`, `hyd pressure2`, `hyd pressure3`, `hyd pressure4`, `mnf flow`, `pressure setpoint`- there appears to be two distributions here. This could potentially be connected to the type of `brand code` or something else not as easily distinguishable. 
* `bowl setpoint` - half of all the values are around 120
* `carb flow` - most values fall between 3,000 and 4,000 with a large pocket of values at 1,000 as well


